Hamiltonian mechanics is conservation of information entropy

### 2018-12-07 Articles/HamiltonianMechanicsIsConservationOfEntropy/article.tex
Reviewed discussions and conclusions

Hamiltonian mechanics, through Liouville's theorem, has always had an important role in statistical mechanics. What we are showing is that that role is not a coincidence: it is the very defining characteristic. \textbf{Hamiltonian mechanics is a type of statistical mechanics}, one with no probability and deterministic and reversible laws. The fact that entropy can be defined objectively and is conserved through evolution cannot be simply seen as a curiosity.

To us, it does not make sense to start our physical discourse by imposing a state space for point-like particles, that somehow does not require only points but a particular differentiable structure as well. Then, if we are so keen, put distributions on it which turn out, somewhat magically, to have density and entropy defined in a coordinate independent way. It makes a lot more sense to start by saying we have a finite system which we pretend to be made of infinitesimal parts, which we keep dividing as we please and derive the state space of particles as the limit of such process of subdivision. This simultaneously gives the actual mathematical structure and its physical meaning.

Not only is the second approach preferable because with fewer starting points it justifies the state space, the evolution and why that state space is so compatible with statistical mechanics, it also maps better to what we do in classical physics. We study composite systems (planets, cannonballs, beads on a wire, fluids) which we sometime pretend we can recursively divide. None of those objects, in fact, has a single position or momentum: only the center of mass has a single position and momentum but those are precisely the expectation values $\int q \rho(q,p) dq dp$ and $\int p \rho(q,p) dq dp$ over the distribution. In other words, particles are not point-like in classical mechanics. They are infinitesimal cells of phase-space and that is why their evolution is symplectomorphism.

Under this light, asking whether classical mechanics is tenable coincides with asking whether we can really keep dividing our object indefinitely. If one is convinced that that is not realistic then he is already convinced that classical mechanics will not hold in those regimes. And if that approximation fails, it is not that we need an altogether new set of rules: the macroscopic object is still the same. We simply need to change how the subdivision works.

Also note that this view tells us exactly what the mathematical framework of symplectic geometry is supposed to represent. The state space of particles is a manifold because we use real numbers to identify the states. Among all possible choices of state variables, we conveniently choose the ones in which the distribution can be expressed as a density. That is the differentiable structure on the manifold. The tangent and cotangent space are the tools to keep track of the process of infinitesimal subdivision and they are linear spaces because the space of all possible distributions is itself a linear space. Moreover the differentiable manifold has a symplectic structure because the density is invariant under coordinate transformation. Therefore among all possible choices of differentiable state variables we choose the canonical ones, the ones in which the density is expressed in appropriate units. Every requirement is physically motivated. No extra requirements are left.

While the power of this shift can really be appreciated in the context of our broader work, it should be clear how with relative few starting points, elements of relativity (e.g. coordinate invariance), statistical mechanics (e.g. physical distributions), quantum mechanics (e.g. wave number, conjugate quantities) and classical Hamiltonian mechanics (e.g. phase space, Poisson brackets) already emerge and form a coherent picture. In particular note how energy (in the form of the Hamiltonian) and entropy (in the form of information entropy) both come out from the single unassuming idea of invariant distributions. And note how conservation of entropy is related to deterministic and reversible evolution. Also note how deterministic and reversible evolution means that the system is only influenced by itself, which means it is not influenced by anything else, which means it is isolated, which is linked to energy conservation.\footnote{To be precise, conservation of energy is not an invariant property. The fact that a Hamiltonian exists, though, is an invariant property.} Surely this cannot be all coincidence.

The larger point is that physics is one and it cannot be understood piecemeal. Nature does not separate between dynamical systems and thermodynamics, between topologies and $\sigma$-algebras, between classical systems and quantum systems. The boundaries between all these disciplines are as much unfortunate as they are artificial. There is no real understanding of one without the other.

Having a coherent picture across the different areas is not only intellectually more satisfying, but it allows us to switch perspective when the intuition from one fails. As a practical example, we can show how understanding the link between Hamiltonian mechanics and information theory leads very straightforwardly to a classical equivalent for the uncertainty principle.

@@

As we saw, Hamiltonian dynamics not only conserves the energy of all particles, but it also conserves the information entropy of the distribution as a whole. We can imagine that there is going to be some link between the spread of a distribution and the number of bits required to identify an element. So we ask: what is the distribution that minimizes the spread given a certain amount of information entropy? If $I_0$ is the set amount of information entropy and $\sigma_q^2 \sigma_p^2 \equiv \int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp$ is the spread, we can use Lagrange multipliers to find out.
\begin{align*}
L = &\int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp \\
&+ \lambda_1(\int \rho dqdp - 1) \\ &+ \lambda_2(- \int \rho \ln \rho \, dqdp - I_0)\\ 
\delta L = &\int \delta \rho [(q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 + \\ &\lambda_1 - \lambda_2 \ln \rho - \lambda_2 ] dqdp = 0 \\
\lambda_2 \ln \rho = &\lambda_1 - \lambda_2 + (q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\
\rho = &e^{\frac{\lambda_1 - \lambda_2}{\lambda_2}}e^{\frac{(q-\mu_q)^2 \sigma_p^2}{\lambda_2}}e^{\frac{\sigma_q^2 (p-\mu_p)^2}{\lambda_2}}\\
\end{align*}
We solve the multipliers and have:
\begin{align*}
\rho = &\frac{1}{ 2 \pi \sigma_q \sigma_p} e^{-\frac{(q-\mu_q)^2}{2\sigma_q^2}} e^{-\frac{(p-\mu_p)^2}{2\sigma_p^2}} \\
I_0 = &\ln (2\pi\sigma_q\sigma_p) + 1
\end{align*}
The distribution that minimizes the spread, then, is the product of two independent Gaussians. As the entropy is conserved during Hamiltonian evolution the product $\sigma_q^2 \sigma_p^2$ can never be less than the one given by the Gaussian distribution of the same entropy. We have:
\begin{align*}
\sigma_q\sigma_p \geq \exp (I_0 - 1) / 2 \pi 
\end{align*}
Intuitively, Hamiltonian mechanics is deterministic and reversible evolution for a distribution and its elements, and therefore it cannot shrink the distribution more than a set amount. If it did, it would start concentrating the density, meaning that multiple states have part of their density mapped to a single state and the dynamics would not be reversible.

@@

In this paper we have shown how classical Hamiltonian mechanics coincides with conservation of information entropy. The structure of phase space is required to be able to define densities and entropy in a way that is coordinate invariant, and therefore physically meaningful. This clarifies the link between Hamiltonian mechanics, statistical mechanics and deterministic and reversible motion.

In the narrower context, it allows us to understand what physical object is represented by each mathematical one. This can be summarized in table \ref{dictionary}.
\begin{table}[h]
	\centering
	%	\begin{tabular}{p{0.2\textwidth} p{0.1\textwidth} p{0.1\textwidth} p{0.5\textwidth}}
	\begin{tabular}{c p{0.3\textwidth} p{0.5\textwidth} }
		& Name & Meaning\\ 
		\hline 
		& Classical particle & an infinitesimal part of the system (i.e. the limit of recursive division) \\ 
		$\mathcal{S} =\mathsf{T}^*\mathcal{Q}$ & Phase space \newline (Cotangent bundle) & the set of all possible states for particles upon which invariant distributions can be defined \\
		$\xi^a$ & State variables \newline (Unified coordinates) & the set of variables needed to identify the state of a particle \\ 
		$\rho(\xi^a)$ & Density & the amount of material for a given particle state (i.e. in the limit of the recursive division)\\ 
		$I(\rho(\xi^a))$ & Information entropy & the number of bits to identify an element of the distribution to a unitary level of precision\\ 
		$q^i$ & Coordinate & a state variable that also defines a unit \\
		$\mathcal{Q}$ & Coordinate space \newline (Configuration space) & the space charted by all coordinates \\
		$k_i$ & Conjugate coordinate & a state variable that uses the inverse unit of the corresponding coordinate \\
		$\hbar$ & & the unit for measuring the range of possible states within a single degree of freedom \\
		$p_i=\hbar k_i$ & Conjugate momentum & a state variable that together with the corresponding coordinate defines invariant ranges of possible states \\
		$(q^i, p_i)$ & Canonical variables \newline (Canonical coordinates) & a set of variables for which the density is invariant and expressed over units of $\hbar$ per degree of freedom\\ 
		& Canonical transformation & a change in state variables that does not change the density\\ 
	\end{tabular}
	\caption{Dictionary between mathematical and physical objects.}
	\label{dictionary}
\end{table}
It also allows us to see why deterministic and reversible evolution must conserve entropy and why isolated systems must conserve energy. In the broader context, it allows us to make deeper connections among different areas of math and physics, and pushes us to rethink our starting points. In particular, it tells us that classical particles should not be considered point-like objects (upon which entropy and densities cannot be defined), but the limit of recursive subdivisions (i.e. infinitesimal cells of phase space).

These results represent one set of insights from our larger project, Assumptions of Physics, that aims to rederive the standard laws from a handful of physical starting points. We believe that a reorganization of fundamental physics on more physically meaningful conceptual footing and more rigorous mathematical grounds is not only possible, but long overdue.

>>>>

Hamiltonian mechanics, through Liouville's theorem, has always had an important role in statistical mechanics. What we are showing is that that role is not a coincidence: it is the very defining characteristic. \textbf{Hamiltonian mechanics is a type of statistical mechanics}, one with no probability and deterministic and reversible laws.

The usual approach posits point-like particles, with a particular state space and equations of motion. Then, coincidentally, finds this setting convenient for statistical mechanics because it has nice properties. In retrospect, this should sound backwards. We believe the approach presented here is preferable, first of all, because it justifies with a single insight the state space, the evolution and why that state space is so compatible with statistical mechanics.

Second, it maps better to what we do in classical physics. We study composite systems (planets, cannonballs, beads on a wire, fluids) which we sometimes pretend we can recursively divide. None of those objects, in fact, has a single position or momentum: only the center of mass has a single position and momentum but those are precisely the expectation values $\int q \rho(q,p) dq dp$ and $\int p \rho(q,p) dq dp$ over the distribution. Under this light, asking whether classical mechanics is tenable coincides with asking whether we can really keep dividing our object indefinitely. If one is convinced that this is not realistic then one is already convinced that classical mechanics will not hold in those regimes. But if that approximation fails, we do not need an altogether new set of ideas: the macroscopic object is still the same. We simply need to change how the subdivision works.

Third, this new approach is conceptually more compatible with the other physical theories. We have already seen how elements of special relativity (e.g. coordinate invariance, momentum as covector), statistical mechanics (e.g. joint and marginal distributions), quantum mechanics (e.g. wave number) and classical Hamiltonian mechanics (e.g. phase space, Poisson brackets) already emerged naturally. Furthermore, note that in general relativity point-like particles would be singularities that puncture space-time and would not move along a geodesics, while particles as infinitesimal parts would. Note that if the evolution is deterministic and reversible, it means the evolution of the system depends only on itself, which means that the system is isolated. That is: we have an association between isolated system and conservation of energy.\footnote{To be precise, conservation of energy is not an invariant property. The fact that a Hamiltonian exists, though, is an invariant property.} This sounds strikingly close to the first law of thermodynamics.

The larger point is that physics is one and it cannot truly be understood piecemeal. Nature does not separate between dynamical systems and thermodynamics, between topologies and $\sigma$-algebras, between classical systems and quantum systems. The boundaries between all these disciplines are as much unfortunate as they are artificial. There is no real understanding of one without the other.

Having a coherent set of ideas, a coherent picture across the different areas is not only intellectually more satisfying, but it allows us to switch perspective when the intuition from one fails. As a practical example, we can show how understanding the link between Hamiltonian mechanics and information theory leads to a classical analogue for the uncertainty principle.

@@

As we saw, Hamiltonian dynamics not only conserves the energy of individual particles, but it also conserves the information entropy of the distribution as a whole. Intuitively, we can see that there is some link between the spread of a distribution and the number of bits required to identify an element. For example, if we have two uniform distributions with different ranges, then Alice will need to give Bob more information to identify the value from the bigger range. So we ask: what is the distribution that minimizes the spread given a certain amount of information entropy? If $I_0$ is the set amount of information entropy and $\sigma_q^2 \sigma_p^2 \equiv \int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp$ is the spread, we can use Lagrange multipliers to find out.
\begin{align*}
L = &\int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp \\
&+ \lambda_1(\int \rho dqdp - 1) \\ &+ \lambda_2(- \int \rho \ln \rho \, dqdp - I_0)\\ 
\delta L = &\int \delta \rho [(q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 + \\ &\lambda_1 - \lambda_2 \ln \rho - \lambda_2 ] dqdp = 0 \\
\lambda_2 \ln \rho = &\lambda_1 - \lambda_2 + (q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\
\rho = &e^{\frac{\lambda_1 - \lambda_2}{\lambda_2}}e^{\frac{(q-\mu_q)^2 \sigma_p^2}{\lambda_2}}e^{\frac{\sigma_q^2 (p-\mu_p)^2}{\lambda_2}}\\
\end{align*}
We solve the multipliers and have:
\begin{align*}
\rho = &\frac{1}{ 2 \pi \sigma_q \sigma_p} e^{-\frac{(q-\mu_q)^2}{2\sigma_q^2}} e^{-\frac{(p-\mu_p)^2}{2\sigma_p^2}} \\
I_0 = &\ln (2\pi\sigma_q\sigma_p) + 1
\end{align*}
The distribution that minimizes the spread, then, is the product of two independent Gaussians. As the entropy is conserved during Hamiltonian evolution the product $\sigma_q^2 \sigma_p^2$ can never be less than the one given by the Gaussian distribution of the same entropy. We have:
\begin{align*}
\sigma_q\sigma_p \geq \exp (I_0 - 1) / 2 \pi 
\end{align*}
Intuitively, Hamiltonian mechanics is deterministic and reversible evolution for a distribution and its elements, and therefore it cannot shrink the distribution more than a set amount. If it did, it would start concentrating the density, meaning that multiple states would be mapped to a single state and the dynamics would not be reversible. In this form $I_0$ is not a fundamental constant but a parameter that depends upon the initial condition. While this is already striking, \cite{AoPPhy1} shows a simple argument to turn this into something more.

Suppose that we have a distribution where the particles move non-derministically. That is, their motion is partly caused by some other agent in a way that can be characterized by random motion. Now suppose our distribution is at an equilibrium with that random motion. Then the particles change state randomly but the overall distribution remains the same. That is, the random motion simply reshuffles the elements. In this case the minimum achievable for the information entropy of the distribution depends on the random motion itself. The stronger it is, the more spread the distribution has to be, the higher the minimum threshold for the information entropy.

Now suppose that source of chaotic motion is something intrinsic to a physical system, something it cannot be avoided. Maybe it is caused by interaction with the environment, from which the system can never be fully isolated. Or maybe it is caused by some internal motion, which can never be fully accessible. Or maybe it is caused by some background process with space itself, since the vacuum is not just empty space. The point here is not to identify the source of this non-deterministic motion: the point is to show that assuming one exists does not sound unreasonable.

If one buys into the argument, one can posit that there exists an $I_\emptyset$ associated with this chaotic process that acts as the theoretical limit for the information entropy of our distributions. The farther the information entropy of our distribution is from that value, the more we can assume that the parts of the distribution evolve deterministically. The closer it is, the more our assumptions breaks down. This is what \cite{AoPPhy1} refers to as the classical uncertainty principle.

\subsection*{A dictionary}

We mentioned several time how this approach allows us to give precise physical meaning to the different mathematical objects. We leave in table \ref{dictionary} a summary of these connections in form of a math-to-physics dictionary.
\begin{table}[h!]
	\centering
	%	\begin{tabular}{p{0.2\textwidth} p{0.1\textwidth} p{0.1\textwidth} p{0.5\textwidth}}
	\begin{tabular}{c p{0.3\textwidth} p{0.5\textwidth} }
		& Name & Meaning\\ 
		\hline 
		& Classical particle & an infinitesimal part of the system (i.e. the limit of recursive division) \\ 
		$\mathcal{S} =\mathsf{T}^*\mathcal{Q}$ & Phase space \newline (Cotangent bundle) & the set of all possible states for particles upon which invariant distributions can be defined \\
		$\xi^a$ & State variables \newline (Unified coordinates) & the set of variables needed to identify the state of a particle \\ 
		$\rho(\xi^a)$ & Density & the amount of material for a given particle state (i.e. in the limit of the recursive division)\\ 
		$I(\rho(\xi^a))$ & Information entropy & the number of bits to identify an element of the distribution to a unitary level of precision\\ 
		$q^i$ & Coordinate & a state variable that also defines a unit \\
		$\mathcal{Q}$ & Coordinate space \newline (Configuration space) & the space charted by all coordinates \\
		$k_i$ & Conjugate coordinate & a state variable that uses the inverse unit of the corresponding coordinate \\
		$\hbar$ & & the unit for measuring the range of possible states within a single degree of freedom \\
		$p_i=\hbar k_i$ & Conjugate momentum & a state variable that together with the corresponding coordinate defines invariant ranges of possible states \\
		$(q^i, p_i)$ & Canonical variables \newline (Canonical coordinates) & a set of variables for which the density is invariant and expressed over units of $\hbar$ per degree of freedom\\ 
		& Canonical transformation & a change in state variables that does not change the density\\ 
	\end{tabular}
	\caption{Dictionary between mathematical and physical objects.}
	\label{dictionary}
\end{table}

We can also better understand what physical requirement each mathematical structure is used for. Manifolds are used to represent a set of physical objects that can be identified by $n$ continuous variables.  A differentiable structure on a manifold is used to define densities. Non-differentiable transformations, in fact, do not define a Jacobian and therefore cannot express the density in the new variables. A symplectic manifold is used to define coordinate invariant densities. Non-canonical transformations, in fact, have non unitary Jacobians and therefore the density would change the value of the density.

@@

In this paper we have shown how classical Hamiltonian mechanics coincides with conservation of information entropy. The structure of phase space is required to be able to define densities and information entropy in a way that is coordinate invariant, and therefore physically meaningful. The equations of motion and the existence of the Hamiltonian are required to conserve densities and information entropy over time. This clarifies the link between Hamiltonian mechanics, statistical mechanics and deterministic and reversible motion.

In the narrower context, it allows us to understand what physical object is represented by each mathematical one. In the broader context, it allows us to make deeper connections among different areas of math and physics, and pushes us to rethink our starting points. In particular, it tells us that classical particles should not be considered point-like objects (upon which entropy and densities cannot be defined), but the limit of recursive subdivisions (i.e. infinitesimal cells of phase space).

%These results represent one set of insights from our larger project, Assumptions of Physics, that aims to rederive the standard laws from a handful of physical starting points.

This approach may point to a reorganization of fundamental physics on a more physically rigorous conceptual footing and more meaningful mathematical grounds which we believe is not only possible, but long overdue.
